### 2020-07-27 manuscript.tex
Refactor

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Smarter dating and faster proposals: revisiting the phylogenetic relaxed clock model} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jordan Douglas\textsuperscript{1,2*},
Rong Zhang\textsuperscript{1,2},
Alexei J. Drummond\textsuperscript{1,2,3},
Remco Bouckaert\textsuperscript{1,2}
\\
\bigskip
\textbf{1} Centre for Computational Evolution,  University of Auckland, Auckland, New Zealand\\
\textbf{2} School of Computer Science, University of Auckland, Auckland, New Zealand\\
\textbf{3} School of Biological Sciences, University of Auckland, Auckland, New Zealand
\\
\bigskip

@@

With the increasing availabilty of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational runtime but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of clock models, smart tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

While a range of advanced operators and other MCMC optimisation methods have arisen over the years, there has yet to be a widescale performance benchmarking of such methods as applied to the relaxed clock model. 
In this article, we systematically evaluate how the relaxed clock model can benefit from i) different substitution rate parameterisations,  ii) the use of Bactrian proposal kernels \cite{yang2013searching}, iii) tree operators which account for correlations between substitution rates and times, and iv) adaptive multivariate operators \cite{baele2017adaptive}.
The discussed methods are implemented in and compared using BEAST2 \cite{bouckaert2019beast}.

@@

% Define tau

@@

for substitution rate standard deviation $\sigma$ and other model parameters $\theta$. $\vec{\mathcal{R}}^{\,}$ is a vector of abstracted substitution rates, which is transformed into real rates by $r(\vec{\mathcal{R}}^{\,})$. Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

@@

\subsection*{Substitution rate parameterisations}
\label{sect:rateparams}

@@

This method has been widely used in BEAST and BEAST2 analyses \cite{drummond2006relaxed}. 
However, the recently developed constant distance operators  -- which are incompatible with the \textit{cat} parameterisation -- yield an increase in mixing rate under \textit{real} by up to an order of magnitude over that of \textit{cat} \cite{zhang2020improving}.

@@

Transforming these quantiles into rates invokes the i-cdf of the underlying \textit{real} clock model distribution. Thus, while this approach has clear similarities with \textit{cat}, the domain of rates here is continuous (as opposed to being confined to a discrete number of bins) and is therefore compatible with constant distance operators \cite{zhang2020improving}.

A potential disadvantage of the \textit{quant} method would be the computational requirements of continuously evaluating the i-cdf.
As the number of quantiles grows with the taxon count $N$, this drawback would be at its worst on large trees.
Hence, rather than evaluating the exact i-cdf $F^{-1}$, an approximation $\hat{F}^{-1}$ will be used instead:

\begin{align}
r(\mathcal{R}_i) =& \; \hat{F}^{-1}(\mathcal{R}_i).
\end{align}

In this article we have extended \textit{quant} through a linear piecewise approximation of the i-cdf. As the piecewise approximation is linear, evaluating the derivatives $\frac{\partial}{\partial \mathcal{R}_i} \hat{F}^{-1}(\mathcal{R}_i) =  D \hat{F}^{-1}(\mathcal{R}_i)$ and $\frac{\partial}{\partial r_i} \hat{F}(r_i) = D \hat{F}(r_i)$  -- which are required for computing the Hastings ratios -- is trivial. The approximation is comprised of $n$ pieces (where $n$ is fixed) and upper and lower rate boundaries $r_\text{min}$ and $r_\text{max}$. The approximation is displayed in \textbf{Fig \ref{fig:rateparams}} and further detailed in \textbf{\nameref{sect:S1Appendix}}.

Zhang and Drummond 2020 introduced several tree operators for the \textit{real} parameterisation -- including \texttt{Constant Distance}, \texttt{Simple Distance}, and \texttt{Small Pulley}. In this project, we extended these three operators so that they are compatible with the \textit{quant} parameterisation. These are presented in \textbf{\nameref{sect:S1Appendix}}.

\subsection*{Bactrian proposal kernel} \label{sect:randomwalks}

@@

In this article we compare the performance of uniform and Bactrian proposal kernels in the clock model. Two Bactrian distributions are compared($m=0.95$ and $m=0.98$). The clock model operators which these proposal kernels apply to are described in \textbf{Table \ref{table:kernels}}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l|} 
 \hline
 & Operator & Proposal & Parameters  \\
 1 & \texttt{Random walk} operator & $x^\prime \leftarrow x + s\Sigma$ & $\sigma, r, q$ \\
 2 & \texttt{Scale} operator & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma, r$  \\
 3 & \texttt{Interval} operator & $\begin{array} {rl} y &\leftarrow \frac{u - x}{x - l} \times e^{s\Sigma} \\ x^\prime &\leftarrow \frac{u + l * y}{y + 1}  \end{array}$ & $q \;\; (l=0,u=1)$  \\
 4 & \texttt{Constant distance} operator & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\

 \hline
\end{tabular}
\caption{Summary of proposal kernels $q(x^\prime|x)$ of clock model operators.
\texttt{Random walk}, \texttt{Scale}, and \texttt{Interval} are standard BEAST2 operators.
 In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution (distributions are normalised so that they have a mean of 0 and a variance of 1). The scale size $s$ is tunable. The proposal kernel may apply to node heights $t$, clock standard deviation $\sigma$, clock rates $r$ (\textit{real} only), and clock rate quantiles $q$ (\textit{quant} only). The Scale operator acts on parameters with non-negative domains. The Interval operator proposes values which respect the domain of the parameter i.e. $l < x, x^\prime < u$. }
\label{table:kernels}
\end{table}

@@

\newpage
% REMEMBER to define NNI and SPR acronyms in introduction preferably
\subsection*{Narrow Exchange Rate} \label{sect:NER}

The \texttt{Narrow Exchange} operator \cite{drummond2002estimating}, widely used in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST2 \cite{bouckaert2019beast}, is similar to NNI, and works as follows (\textbf{Fig. \ref{fig:narrowexchange}}):

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{Narrow Exchange} and \texttt{Narrow Exchange Rate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axis corresponds to node height $t$. In the bottom figure, branch rates $r$ are indicated by line thickness. In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.}
\label{fig:narrowexchange}
\end{figure}

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI and SPR) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). If \texttt{Narrow Exchange} was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further. This may in turn permit proposing a new node height $t_D$ and therefore changing branch (time) lengths.

@@

Here, we present the \texttt{Narrow Exchange Rate} (NER) operator. Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. In addition to the modest topological change applied by \texttt{Narrow Exchange}, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. While NER does not alter $t_D$ (i.e. ${t_D}^\prime \leftarrow t_D$), we also consider NERw - a special case of the NER operator which embarks $t_D$ on a random walk:

@@

for random walk step size $s\Sigma$ where $s$ is a tunable scaler parameter and $\Sigma$ is drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the \texttt{Constant Distance} operator, new rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

@@

The total space of NER operators is comprised of all possible subsets of distance constraints (i.e. $\{\},\{\mathcal{D}_{AB}\}, \{\mathcal{D}_{AC}\}, \dotso , \{\mathcal{D}_{AB}, \mathcal{D}_{AC}, \mathcal{D}_{AE}, \mathcal{D}_{BC}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$) which are solvable. The simplest NER -- the null operator denoted by NER$\{\}$ -- does not satisfy any distance constraints. This is equivalent to \texttt{Narrow Exchange}.

@@

\subsection*{A guided adaptive leaf rate operator}
\label{AVMVN_sect}

@@

In this article we consider application of the AVMVN kernel to the branch rates of leaf nodes. This operator is not readily applicable to internal node branch rates due to their dependencies on tree topology.

@@

In \nameref{sect:results}, we evaluate this operator for its ability to estimate leaf rates. As the size of the covariance matrix $\Sigma_N$ grows with the number of taxa $N$, AVMVN is hypothesised to work well on small trees but become less efficient with larger taxon sets.

\section*{Results} \label{sect:results}

@@

To avoid a cross-product explosion, the four targets for clock model improvement (\textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:randomwalks}}, \textbf{\nameref{sect:NER}}, and \textbf{\nameref{AVMVN_sect}}) are evaluated sequentially, in the order presented in this paper. The setting which is considered to be the best in each step is then incorporated into the following step. This protocol and its outcomes are summarised in \textbf{Fig. \ref{fig:tournament}}.

@@

\subsection*{Evaluating constant distance operators}

\subsection*{Comparison of rate parameterisations}

We compared the three rate parameterisations described in \textbf{\nameref{sect:rateparams}}. All three settings use the standard BEAST2 clock model operators from Drummond et al. 2006 \cite{drummond2006relaxed}. \textit{real} and \textit{quant} additionally use the constant-distance tree operators \cite{zhang2020improving}. To determine whether the difference in performance between \textit{real}/\textit{quant} versus \textit{cat} is because of the constant-distance tree operators or the parameterisation itself, we also included benchmarked two additional settings: \textit{real 2006} and \textit{quant 2006}, which do not used the constant-distance operators. These five settings are validated in \textbf{\nameref{sect:WCSS_appendix}}.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Adaptive dating and faster proposals: revisiting the phylogenetic relaxed clock model} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jordan Douglas\textsuperscript{1,2*},
Rong Zhang\textsuperscript{1,2},
Alexei J. Drummond\textsuperscript{1,2,3},
Remco Bouckaert\textsuperscript{1,2}
\\
\bigskip
\textbf{1} Centre for Computational Evolution,  University of Auckland, Auckland, New Zealand\\
\textbf{2} School of Computer Science, University of Auckland, Auckland, New Zealand\\
\textbf{3} School of Biological Sciences, University of Auckland, Auckland, New Zealand
\\
\bigskip

@@

With the increasing availabilty of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational runtime but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels including the Bactrian and mirror kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of clock models, smart tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

While a range of advanced operators and other MCMC optimisation methods have arisen over the years, there has yet to be a widescale performance benchmarking of such methods as applied to the relaxed clock model. 
In this article, we systematically evaluate how the relaxed clock model can benefit from i) adaptive operator weighting, ii) different substitution rate parameterisations,  iii) the use of Bactrian proposal kernels \cite{yang2013searching}, iv) tree operators which account for correlations between substitution rates and times, and v) adaptive multivariate operators \cite{baele2017adaptive}.
The discussed methods are implemented in and compared using BEAST2 \cite{bouckaert2019beast}.

@@

@@

where $\sigma$ is the clock model standard deviation, $p(\mathcal{T}|\theta)$ is the tree prior, and $\theta$ describes other unspecified parameters. 
The tree likelihood $p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \theta)$ is computed using Felsenstein's tree-peeling algorithm \cite{felsenstein1981evolutionary}, where
$\vec{\mathcal{R}}^{\,}$ is a vector of abstracted substitution rates and transformed into real rates by function $r(\vec{\mathcal{R}}^{\,})$. Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

@@

\clearpage
\subsection*{Substitution rate parameterisations}
\label{sect:rateparams}

@@

This parameterisation has been widely used in BEAST and BEAST2 analyses \cite{drummond2006relaxed}. 
However, the recently developed constant distance operators  -- which are incompatible with the \textit{cat} parameterisation -- can yield an increase in mixing rate under \textit{real} by up to an order of magnitude over that of \textit{cat}, depending on the dataset \cite{zhang2020improving}.

@@

Transforming these quantiles into rates invokes the i-cdf of the underlying \textit{real} clock model distribution:

\begin{align}
r(\mathcal{R}_i) =& \; F^{-1}(\mathcal{R}_i).
\end{align}

While this approach has clear similarities with \textit{cat}, the domain of rates here is continuous (as opposed to being confined to a discrete number of bins).
In this project we extended the family of constant distance operators  \cite{zhang2020improving} so they are compatible with \textit{quant}.
These are presented in \textbf{\nameref{sect:S1Appendix}}.

%A potential disadvantage of the \textit{quant} method would be the computational requirements of continuously evaluating the i-cdf.
%As the number of quantiles grows with the taxon count $N$, this drawback would be at its worst on large trees.
%Hence, rather than evaluating the exact i-cdf $F^{-1}$, an approximation $\hat{F}^{-1}$ will be used instead:

%\begin{align}
%r(\mathcal{R}_i) =& \; \hat{F}^{-1}(\mathcal{R}_i).
%\end{align}

%In this article we have extended \textit{quant} through a linear piecewise approximation of the i-cdf. As the piecewise approximation is linear, evaluating the derivatives $\frac{\partial}{\partial \mathcal{R}_i} \hat{F}^{-1}(\mathcal{R}_i) =  D \hat{F}^{-1}(\mathcal{R}_i)$ and $\frac{\partial}{\partial r_i} \hat{F}(r_i) = D \hat{F}(r_i)$  -- which are required for computing the Hastings ratios -- is trivial. The approximation is comprised of $n$ pieces (where $n$ is fixed) and upper and lower rate boundaries $r_\text{min}$ and $r_\text{max}$. The approximation is displayed in \textbf{Fig \ref{fig:rateparams}} and further detailed in \textbf{\nameref{sect:S1Appendix}}.

%Zhang and Drummond 2020 introduced several tree operators for the \textit{real} parameterisation -- including \texttt{Constant Distance}, \texttt{Simple Distance}, and \texttt{Small Pulley}. In this project, we extended these three operators so that they are compatible with the \textit{quant} parameterisation. These are presented in \textbf{\nameref{sect:S1Appendix}}.

\clearpage
\subsection*{Adaptive operator weighting}
\label{sect:adaptiveSampling}

The weight of an operator determines the probability of the operator being selected at each step during MCMC.
In BEAST2, operators can have a tunable parameter $s$ which determines the step size of the operator \cite{bouckaert2019beast}. 
Although $s$ is learned throughout the MCMC chain, the operator weights themselves are typically held constant.
Pre-existing BEAST2 clock model operators (i.e. those which generate proposals for either $\mathcal{R}$ or $\sigma$) are summarised in \textbf{Table \ref{table:kernels}}, and further operators are introduced throughout the paper.

% Section intro
% Give some examples of this operator working better on this dataset but not that one, where both operators apply to the same parameter eg. Constant Distance? Any others?
% Bring up theme that this can be related to signal. Sampling from the prior may give better results when there is poor signal, whereas random-walk based operators may do better in the other case. Any general theoretical examples of this?

\begin{table}[h!]
\centering
\begin{tabular}{|l| l p{4cm} l l|} 
 \hline
 & Operator & Description & Parameters & Parameterisations  \\
  \hline
 1 & \texttt{RandomWalk} & Moves a single element by a tunable amount. & $\mathcal{R}, \sigma$ & \textit{cat}, \textit{real}, \textit{quant} \\
  \hline
 2 & \texttt{Scale} & Applies \texttt{RandomWalk} on the log-transformation (suitable for parameters with positive domains). & $\mathcal{R}, \sigma$ & \textit{real}, \textit{quant}  \\
  \hline
 3 & \texttt{Interval} & Applies \texttt{RandomWalk} on the logit-transformation (suitable for parameters with upper and lower limits). & $\mathcal{R}$ & \textit{quant}  \\
  \hline
   4 & \texttt{Swap} & Swaps two random elements in the vector \cite{drummond2006relaxed}. & $\mathcal{R}$  & \textit{cat}, \textit{real}, \textit{quant}  \\
 \hline
  5 & \texttt{Uniform} & Resamples one element in the vector from a uniform distribution. & $\mathcal{R}$  & \textit{cat}, \textit{quant}  \\
 \hline
 6 & \texttt{ConstantDistance} & Adjusts an internal node and recalculates all incident branch rates such that the genetic distances remain constant \cite{zhang2020improving}.  & $\mathcal{R}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
 7 & \texttt{SimpleDistance} & Applies \texttt{SimpleDistance} to the root node \cite{zhang2020improving}.  & $\mathcal{R}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
 8 & \texttt{SmallPulley} & Proposes new branch rates incident to the root such that their combined genetic distance is constant  \cite{zhang2020improving}.  & $\mathcal{R}$ & \textit{real}, \textit{quant} \\
 \hline
 9 & \texttt{FastClockScale} & Applies \texttt{Scale} to $\sigma$ and then recomputes all rates $\mathcal{R}$ such that their quantiles under the prior $p(\mathcal{R}|\sigma)$ are constant \cite{zhang2020improving}.  & $\mathcal{R}$ & \textit{real} \\
 \hline
\end{tabular}
\caption{Summary of pre-existing BEAST2 operators which apply to either branch rates $\mathcal{R}$ or the clock standard deviation $\sigma$, and the substitution rate parameterisation they apply to.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} also adjust tree node heights. }
\label{table:kernels}
\end{table}

It is not always clear which mixture of operators is best for a given dataset.
In this article we introduce \texttt{AdaptiveOperatorSampler} -- a meta-operator which learns operator weights during MCMC and samples operators according to these weights.
The meta-operator undergoes three phases.
In the first phase (burn-in), \texttt{AdaptiveOperatorSampler} samples from its set of sub-operators uniformly at random.
In the second phase (learn-in), the meta-operator continues to sample operators uniformly at random however it begins learning several terms detailed below.
In its final phase, \texttt{AdaptiveOperatorSampler} samples operators (denoted by $\omega$) using the following distribution:

\begin{align}
\label{eqn:adaptiveSampler}
	p(\omega_i) \propto \begin{cases} 1 & \text{ with probability } \Omega \\ \frac{1}{\mathbb{T}(\omega_i)} \sum\limits_{p \in \text{parameters}} \frac{1}{\sigma^2_p} \Big[ \sum\limits_{x \in \text{accepts}_i}  (x_p - x_p^\prime)^2 \Big] & \text{ with probability } 1-\Omega \end{cases}.
\end{align}

where $\Omega = 0.01$ allows any sub-operator to be sampled regardless of its performance.
The remaining terms are trained during the second and third phases: the cumulative computational runtime spent on each operator $\mathbb{T}(\omega_i)$, the sample variance $\sigma^2_p$ of each parameter $p$, and the sum-of-squares $\sum\limits_{x \in \text{accepts}_i} (x_p - x_p^\prime)^2$, where $x_p$ and $x_p^\prime$ are the values of $p$ before and after each acceptance of a proposal made by $\omega_i$.

%The terms which are learned during the second and third phases are: ,

Under Equation \ref{eqn:adaptiveSampler}, operators which effect larger changes on the parameters of interest, in shorter runtime, are sampled with greater probabilities. 
Division of the sum-of-squares term by the parameter variance $\sigma^2_p$ enables comparison between different parameters.

We also introduce the \texttt{SampleFromPrior($\vec{x}^{\,}$)} operator.
This operator resamples $\psi$ randomly selected elements within vector $\vec{x}^{\,}$ from their prior distributions, where $\psi \sim \text{Binomial}(n=|\vec{x}^{\,}|, p=\frac{s}{|\vec{x}^{\,}|})$ for tunable parameter $s$.

We hypothesis that datasets with strong signal (or large $L$) will mix best when the more precise and meticulous kind of operator is employed, such as those informed by correlations within the posterior distribution e.g. the family of constant distance operators \cite{zhang2020improving}.
Whereas, datasets which contain very poor signal (or small $L$) are likely to mix best when there is more weight placed on bold operators such as \texttt{SampleFromPrior}, \texttt{Swap}, or \texttt{Uniform} which sample each branch rate independently of its current estimate.
\texttt{AdaptiveOperatorSampler} should learn the combination of weights behind these classes of operators best suited for the dataset.

In this article we apply three instances of the \texttt{AdaptiveOperatorSampler} meta-operator to the \textit{real} and \textit{quant} parameterisations. 
These are summarised in \textbf{Table \ref{table:adaptiveSampling}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l p{4cm}|} 
 \hline
 Meta-operator & Operators \\
\hline
 \multirow{4}{*}{\texttt{AdaptiveOperatorSampler($\sigma$)}} & \texttt{FastClockScale($\sigma, \mathcal{R}$)}$^{*}$ \\ 
 & \texttt{RandomWalk($\sigma$)} \\
 & \texttt{Scale($\sigma$)} \\
 & \texttt{SampleFromPrior($\sigma$)} \\
 \hline
  \multirow{6}{*}{\texttt{AdaptiveOperatorSampler($\mathcal{R}$)}} & \texttt{ConstantDistance($\mathcal{R}, \mathcal{T}$)} \\ 
& \texttt{RandomWalk($\mathcal{R}$)} \\
& \texttt{Scale($\mathcal{R}$)}$^{*}$ \\
& \texttt{Interval($\mathcal{R}$)}$^{**}$  \\
& \texttt{Swap($\mathcal{R}$)} \\
& \texttt{SampleFromPrior($\mathcal{R}$)} \\
 \hline
   \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(root)}} & \texttt{SimpleDistance($\mathcal{R}, \mathcal{T}$)} \\ 
&  \texttt{SmallPulley($\mathcal{R}, t$)}\\
 \hline
\end{tabular}
\caption{Summary of the three instances of \texttt{AdaptiveOperatorSampler} used under the \textit{real} and \textit{quant} parameterisations.  $^{*}$\textit{real} only. $^{**}$\textit{quant} only. }
\label{table:adaptiveSampling}
\end{table}

%\texttt{AdaptiveOperatorSampler($\sigma$)} samples the clock standard deviation, 
%\texttt{AdaptiveOperatorSampler($\mathcal{R}$)} samples branch rates and internal node heights,
%and \texttt{AdaptiveOperatorSampler(root)} samples

\clearpage
\subsection*{Bactrian proposal kernel} \label{sect:randomwalks}

@@

In this article we compare the performance of uniform and Bactrian proposal kernels in the clock model. Two Bactrian distributions are compared($m=0.95$ and $m=0.98$). The clock model operators which these proposal kernels apply to are described in \textbf{Table \ref{table:bactriankernels}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l| p{3cm} l l|} 
 \hline
 & Operator(s) & Proposal & Parameter $x$   \\
  \hline
 1 & \texttt{RandomWalk}  & $x^\prime \leftarrow x + s\Sigma$ & $\mathcal{R}, \sigma$  \\
  \hline
 2 & \texttt{Scale} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\mathcal{R}, \sigma$   \\
  \hline
 3 & \texttt{Interval} & $\begin{array} {rl} &y \leftarrow \frac{1 - x}{x} \times e^{s\Sigma} \\ &x^\prime \leftarrow \frac{y}{y + 1}  \end{array}$ & $\mathcal{R}$  \\
  \hline
 4 & \texttt{ConstantDistance} \texttt{SimpleDistance} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\
 \hline
 5 & \texttt{SmallPulley} & $x^\prime \leftarrow x + s\Sigma$ & $\mathcal{R}$  \\
 \hline
6 & \texttt{FastClockScale}  & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma$   \\
 \hline
\end{tabular}
\caption{Proposal kernels $q(x^\prime|x)$ of clock model operators.
 In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution (distributions are normalised so that they have a mean of 0 and a variance of 1). 
 The scale size $s$ is tunable.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} propose tree heights $t$.
  The \texttt{Interval} operator applies to rate quantiles and respects its domain i.e. $0 < x, x^\prime < 1$. }
\label{table:bactriankernels}
\end{table}

@@

@@

\clearpage
% REMEMBER to define NNI and SPR acronyms in introduction preferably
\subsection*{Narrow Exchange Rate} \label{sect:NER}

The \texttt{NarrowExchange} operator \cite{drummond2002estimating}, used widely in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST2 \cite{bouckaert2019beast}, is similar to nearest neighbour interchange (NNI), and works as follows (\textbf{Fig. \ref{fig:narrowexchange}}):

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{NarrowExchange} and \texttt{NarrowExchangeRate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axis corresponds to node height $t$. In the bottom figure, branch rates $r$ are indicated by line thickness. In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.}
\label{fig:narrowexchange}
\end{figure}

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). 
If \texttt{NarrowExchange} could be adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further. 
This may in turn permit proposing a new node height $t_D$ and therefore changing branch (time) lengths.

@@

Here, we present the \texttt{NarrowExchangeRate} (NER) operator. 
Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. 
In addition to the modest topological change applied by \texttt{NarrowExchange}, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. While NER does not alter $t_D$ (i.e. ${t_D}^\prime \leftarrow t_D$), we also consider NERw - a special case of the NER operator which embarks $t_D$ on a random walk:

@@

for random walk step size $s\Sigma$ where $s$ is a tunable scaler parameter and $\Sigma$ is drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the \texttt{ConstantDistance} operator, new rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. 
Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

@@

The total space of NER operators is comprised of all possible subsets of distance constraints (i.e. $\{\},\{\mathcal{D}_{AB}\}, \{\mathcal{D}_{AC}\}, \dotso , \{\mathcal{D}_{AB}, \mathcal{D}_{AC}, \mathcal{D}_{AE}, \mathcal{D}_{BC}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$) which are solvable. 
The simplest NER -- the null operator denoted by NER$\{\}$ -- does not satisfy any distance constraints. 
This is equivalent to \texttt{NarrowExchange}.

@@

\clearpage
\subsection*{A guided adaptive leaf rate operator}
\label{AVMVN_sect}

@@

In this article we consider application of the AVMVN kernel to the branch rates of leaf nodes. 
This operator, referred to as \texttt{LeafAVMVN}, is not readily applicable to internal node branch rates due to their dependencies on tree topology.

@@

In \nameref{sect:results}, we evaluate the \texttt{LeafAVMVN} operator for its ability to estimate leaf rates. 
As the size of the covariance matrix $\Sigma_N$ grows with the number of taxa $N$, AVMVN is hypothesised to work well on small trees but become less efficient with larger taxon sets.

\clearpage
\section*{Results} \label{sect:results}

Throughout \nameref{sect:models}, we introduced four new operators which are applied to the relaxed clock. 
These operators are summarised in \textbf{Table \ref{table:newOperators}} and are applied in conjunction with the pre-existing clock model operators presented in \textbf{Table \ref{table:kernels}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l| l p{4cm} l|} 
 \hline
 & Operator & Description & Parameters  \\
  \hline
1 & \texttt{AdaptiveOperatorSampler} & Samples sub-operators proportionally to their weights, which are learned (see \nameref{sect:adaptiveSampling}). & $\mathcal{R}, \sigma, \mathcal{T}$ \\
  \hline
2 & \texttt{SampleFromPrior} & Resamples a random number of elements from their prior (see \nameref{sect:adaptiveSampling}). & $\mathcal{R}, \sigma$ \\
  \hline
3 & \texttt{NarrowExchangeRate} & Moves a branch and recomputes branch rates so that genetic distances are constant (see \nameref{sect:NER}). & $\mathcal{R}, \mathcal{T}$\\
  \hline
4 & \texttt{LeafAVMVN}  & Proposals new rates for all leaves in one move (see \nameref{AVMVN_sect}) \cite{baele2017adaptive}. & $\mathcal{R}$ \\
  \hline
\end{tabular}
\caption{Summary of clock model operators introduced throughout this article.}
\label{table:newOperators}
\end{table}

@@

To avoid a cross-product explosion, the five targets for clock model improvement are evaluated sequentially in the following order: \textbf{\nameref{sect:adaptiveSampling}} \textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:randomwalks}}, \textbf{\nameref{sect:NER}}, and \textbf{\nameref{AVMVN_sect}}.

 The setting which is considered to be the best in each step is then incorporated into the following step. This protocol and its outcomes are summarised in \textbf{Fig. \ref{fig:tournament}}.

@@

\clearpage
\subsection*{Round 1: A simple operator weight learning algorithm can improve performance}

We compared the cons, nocons, and adapt operator configurations for the \textit{real} and \textit{quant} parameterisations (\textbf{Table \ref{table:round1Results}}).
The nocons configuration contained the standard BEAST2 operator mix, while cons contained the operator mix used by Zhang and Drummond 2020 \cite{zhang2020improving}. 
The adapt mix combines the above operators, as well as the rudimentary \texttt{SampleFromPrior} operator, and learns the weights of each operator using  \texttt{AdaptiveOperatorSampler}.
